**The status of the microbial census: an update**

### 2015-11-18 Schloss_Census2_mBio_2015.Rmd
Start filling in Results section

inline_hook <- function(x){
	print(x)
	if(is.numeric(x)){
		if(abs(x - round(x)) < .Machine$double.eps^0.5){
			paste(format(x,big.mark=',', digits=0, scientific=FALSE))
		} else {
			paste(format(x,big.mark=',', digits=2, nsmall=2, scientific=FALSE))
		}

@@

year_of_analysis <- "2015"

bact <- read.table(file='data/process/bacteria.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
bact_year_deposited <- gsub("^(\\d\\d\\d\\d)-.*", "\\1", bact$date)

arch <- read.table(file='data/process/archaea.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
arch_year_deposited <- gsub("^(\\d\\d\\d\\d)-.*", "\\1", arch$date)
```
\vspace{35mm}

@@

***Sequence data curation.***
The July 19, 2015 release of the ARB-formatted SILVA small subunit (SSU) reference database (SSU Ref v.123) was downloaded from http://www.arb-silva.de/fileadmin/silva_databases/release_123/ARB_files/SSURef_123_SILVA_19_07_15_opt.arb.tgz [@REF]. This release is based on the EMBL-EBI/ENA Release 123, which was released in March 2015. The SILVA curators identify potential SSU sequences using keyword searches and sequence-based search using RNAmmer (http://www.arb-silva.de/documentation/release-123/). The SILVA curators then screened the 7,168,241 resulting sequences based on a minimum length criteria (<300 nt), number of ambiguous base calls (>2%), length of sequence homopolymers (>2%), presence of vector contamination (>2%), low alignment quality value (<75), and likelihood of being chimeric (Pintail value < 50). Of the remaining sequences, the bacterial reference set retained those bacterial sequences longer than 1,200 nt and the archaeal reference set retained those archaeal sequences longer than 900 nt. The aligned 1,515,024 bacterial and 59,240 archaeal sequences were exported from the database using ARB along with the complete set of metadata. Additional sequence data was included from single-cell genomes available on the Integrated Microbial Genomes (IMG) system [@REF], many of which were recently obtained via the GEBA-MDM effort in Rinke et al. [@REF]. "SCGC” was searched on the IMG database March 12, 2015 to download the bacterial (N=249) and archaeal (N=46) 16S rRNA gene sequences and their associated metadata. The IMG sequences were aligned against the respective SILVA-based reference using mothur [@REF]. The aligned bacterial and archaeal sequence sets from SILVA and IMG were pooled and processed in parallel in mothur. Using mothur, sequences were further screened to remove any sequence with more than 2 ambiguous base calls and trimmed to overlap the same alignment coordinates while maintaining a minimum sequence length of 1,200 and 900 nt for the bacterial and archeael datasets, respectively. The final datasets contained `r nrow(bact)` bacterial and `r nrow(arch)` archaeal 16S rRNA gene sequences. Sequences were assigned to operational taxonomic units (OTUs) with a 3% distance threshold using the average neighbor clustering algorithm.

@@

Issues we sought to solve...

@@

```{r figure4a}
data <- read.table(file="data/mothur/all_bacteria.env_category.groups.rarefaction", header=T)
category <- "\\.plant$" #change the SO to whatever category you want (leave the \\. and $)
plant_columns <- grep(category, colnames(data))
category <- "\\.aerosol$" #change the SO to whatever category you want (leave the \\. and $)
aerosol_columns <- grep(category, colnames(data))
category <- "\\.aquatic$" #change the SO to whatever category you want (leave the \\. and $)
aquatic_columns <- grep(category, colnames(data))
category <- "\\.built$" #change the SO to whatever category you want (leave the \\. and $)
built_columns <- grep(category, colnames(data))
category <- "\\.other$" #change the SO to whatever category you want (leave the \\. and $)
other_columns <- grep(category, colnames(data))
category <- "\\.soil$" #change the SO to whatever category you want (leave the \\. and $)
soil_columns <- grep(category, colnames(data))
category <- "\\.zoonotic$" #change the SO to whatever category you want (leave the \\. and $)
zoonotic_columns <- grep(category, colnames(data))
plant_aerosol_aquatic_built_other_soil_zoonotic_data <- data[,c(1, plant_columns, aerosol_columns, aquatic_columns, built_columns, other_columns, soil_columns, zoonotic_columns)]  #the 1 gets the first column which is numsampled
nrows_to_keep <- which.max(plant_aerosol_aquatic_built_other_soil_zoonotic_data$numsampled[!is.na(plant_aerosol_aquatic_built_other_soil_zoonotic_data[,2])])
plant_aerosol_aquatic_built_other_soil_zoonotic_data <- plant_aerosol_aquatic_built_other_soil_zoonotic_data[1:nrows_to_keep,]
pdf(file="/Users/renegirard7gfy/Desktop/Bac_Uniq_3per.pdf")

@@

```{r figure4b}

>>>>

inline_hook <- function(x){
	print(x)
	if(is.numeric(x)){
		if(abs(x - round(x)) < .Machine$double.eps^0.5){
			paste(format(x,big.mark=',', digits=0, scientific=FALSE))
		} else {
			paste(format(x,big.mark=',', digits=1, nsmall=1, scientific=FALSE))
		}

@@

bact <- read.table(file='data/process/bacteria.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)

arch <- read.table(file='data/process/archaea.v123.metadata', header=T, row.names=1, stringsAsFactors=FALSE)
```
\vspace{35mm}

@@

***Sequence data curation.***
The July 19, 2015 release of the ARB-formatted SILVA small subunit (SSU) reference database (SSU Ref v.123) was downloaded from http://www.arb-silva.de/fileadmin/silva_databases/release_123/ARB_files/SSURef_123_SILVA_19_07_15_opt.arb.tgz [@REF]. This release is based on the EMBL-EBI/ENA Release 123, which was released in March 2015. The SILVA curators identify potential SSU sequences using keyword searches and sequence-based search using RNAmmer (http://www.arb-silva.de/documentation/release-123/). The SILVA curators then screened the 7,168,241 resulting sequences based on a minimum length criteria (<300 nt), number of ambiguous base calls (>2%), length of sequence homopolymers (>2%), presence of vector contamination (>2%), low alignment quality value (<75), and likelihood of being chimeric (Pintail value < 50). Of the remaining sequences, the bacterial reference set retained those bacterial sequences longer than 1,200 nt and the archaeal reference set retained those archaeal sequences longer than 900 nt. The aligned 1,515,024 bacterial and 59,240 archaeal sequences were exported from the database using ARB along with the complete set of metadata. Additional sequence data was included from single-cell genomes available on the Integrated Microbial Genomes (IMG) system [@REF], many of which were recently obtained via the GEBA-MDM effort in Rinke et al. [@REF]. "SCGC” was searched on the IMG database March 12, 2015 to download the bacterial (N=249) and archaeal (N=46) 16S rRNA gene sequences and their associated metadata. The IMG sequences were aligned against the respective SILVA-based reference using mothur [@REF]. The aligned bacterial and archaeal sequence sets from SILVA and IMG were pooled and processed in parallel in mothur. Using mothur, sequences were further screened to remove any sequence with more than 2 ambiguous base calls and trimmed to overlap the same alignment coordinates while maintaining a minimum sequence length of 1,200 and 900 nt for the bacterial and archeael datasets, respectively. The final datasets contained `r nrow(bact)` bacterial and `r nrow(arch)` archaeal 16S rRNA gene sequences. Sequences were assigned to OTUs using the average neighbor clustering algorithm [@REF].

@@

```{r rarefaction}
get_sobs <- function(data){
	rownames(data) <- data[,1]
	data <- data[,-1]
	data <- data[,!grepl('ci', colnames(data))]
	unlist(data[nrow(data),])
}
b_rare <- read.table(file='data/mothur/all_bacteria.filter.unique.precluster.an.rarefaction', header=T)
b_sobs <- get_sobs(b_rare)
b_goods <- 100*(1-sum(table(bact$otu)==1)/nrow(bact))
b_schloss <- 100*(1-sum(table(bact$otu)==1)/b_sobs["X0.03"])

a_rare <- read.table(file='data/mothur/all_archaea.filter.unique.precluster.an.rarefaction', header=T)
a_sobs <- get_sobs(a_rare)
a_goods <- 100*(1-sum(table(arch$otu)==1)/nrow(arch))
a_schloss <- 100*(1-sum(table(arch$otu)==1)/a_sobs["X0.03"])

***The status of the bacterial and archaeal census.***
To assess the field's progress in characterizing the biodiversity of bacteria and archaea we assigned each 16S rRNA gene sequence to OTUs using distance threshold varying between 0 and 20%. Although it is not possible to link a specific taxonomic level (e.g. species, genus, family, etc.) to a specific distance threshold, we selected distances of 0, 3, 5, 10, and 20% because they are widely regarded as representing the range of genetic diversity of the 16S rRNA gene within each domain. By rarefaction, it was clear that the ongoing sampling efforts have started to saturate the number of current OTUs. After sampling `r nrow(bact)` near full-length bacterial 16S rRNA gene sequences we have identified `r b_sobs['unique']`, `r b_sobs['X0.03']`, `r b_sobs['X0.05']`, `r b_sobs['X0.10']`, and `r b_sobs['X0.20']` OTUs at the selected thresholds (Figure 1A, Table 1). Using on the OTUs generated using a 3% threshold, we found that `r b_goods`% of the sequences belonged to OTUs that had been observed more than once. In contrast, only `r b_schloss`% of the OTUs that were observed had been seen more than once. Paralleling the bacterial results, after sampling `r nrow(arch)` archaeal 16S rRNA gene sequences we have identified `r a_sobs['unique']`, `r a_sobs['X0.03']`, `r a_sobs['X0.05']`, `r a_sobs['X0.10']`, and `r a_sobs['X0.20']` OTUs (Figure 1B, Table 1). Using on the OTUs generated using a 3% threshold, we found that `r a_goods`% of the sequences belonged to OTUs that had been observed more than once. In contrast, only `r a_schloss`% of the OTUs that were observed had been seen more than once. The results for the bacterial and archaeal censuses indicate that regardless of the domain, if we continue sampling with the current strategies we will continue to sample OTUs that have already been observed even though a large fraction of OTUs have only been sampled once. Furthermore, considering more than `r 100-min(b_schloss, a_schloss)`% of the OTUs have only been observed once it is likely that an even larger number of OTUs have yet to be sampled for both domains.

* Why aren't we doing a great job?
  - Rate of change in full-length sequences has plateaued
  - New sequences are submitted by a small number of investigators for a limited number of environments

```{r time_course}
  	time_data <- read.table(file="data/process/by_year_analysis.tsv", header=TRUE)

	get_n50_count <- function(year){
		sort(table(bact[bact_year_deposited==year,'submit_author']), decreasing=T)[1:time_data[as.character(year),"b_n50"]]
	}
	dominant_studies <- sum(unlist(lapply(2009:2012, get_n50_count)))

	get_n50_pm_ids <- function(year){
		authors <- names(sort(table(bact[bact_year_deposited==year,'submit_author']), decreasing=T)[1:time_data[as.character(year),"b_n50"]])
		dom_pm_id <- unique(bact[bact$submit_author %in% authors, c('submit_author', 'pubmed_id')])$pubmed_id
		as.character(dom_pm_id[!is.na(dom_pm_id)])
	}
	dominant_titles <- unlist(lapply(2009:2012, get_n50_pm_ids))

	stopifnot(dominant_titles == c('19478181', '22310478', '20668241', '22310478', '22719818', '22832344'))

	require(knitcitations, quiet=TRUE)
	citep(c('10.1038/ismej.2012.79', '10.1371/journal.pone.0026284', '10.1101/gr.131029.111', '10.1073/pnas.1004204107', '10.1126/science.1171700'))

	get_n50_categories <- function(year){
		authors <- names(sort(table(bact[bact_year_deposited==year,'submit_author']), decreasing=T)[1:time_data[as.character(year),"b_n50"]])
		dom_pm_id <- unique(bact[bact$submit_author %in% authors, c('submit_author', 'category')])$category
		as.character(dom_pm_id[!is.na(dom_pm_id)])
	}
	dominant_categories <- unlist(lapply(c(2009:2012,2014), get_n50_categories))

***The dynamics of the bacterial census.***
One explanation for the large number of OTUs that have only been observed once is that with the the broad adoption of highly parallelized sequencing platforms that generate short sequence reads, the rate of full-length sequence generation has declined. In fact, since 2009 the number of new bacterial sequences generated has plateaued at an average of `r mean(time_data[as.character(2009:2014),'b_nseqs'])` sequences per year (Figure 2A). Although this is still an impressive number of sequences, since 2008 the number of new bacterial OTUs has plateaued at an average of `r mean(time_data[as.character(2006:2014),'b_sobs'])` new OTUs per year (Figure 2B). Given the expense of generating full-length sequences using the Sanger sequencing technology and the transition to other platforms, we expected that the large number of sequences were being deposited by a handful of large projects. Indeed, when we counted the number of submissions responsible for depositing 50% of the sequences, we found that with the exception of 2006 and 2013, eight or fewer studies were responsible for depositing the majority of the full-length sequences each year since 2005 (Figure 2C). Between 2009 and 2012, `r sum(time_data[as.character(2009:2012), 'b_nseqs'])` total sequences were submitted and `r sum(time_data[as.character(2009:2012), 'b_n50'])`) submissions from `r length(unique(dominant_titles))` studies were responsible for depositing `r dominant_studies` (`r 100*dominant_studies/sum(time_data[as.character(2009:2012), 'b_nseqs'])`% of all sequences). These studies generated sequences from the human gastrointestinal tract [@Li_2012], human skin [@Kong_2012; @Grice_2009], murine skin [@Grice_2010], and hypersaline microbial mats [@Kirk_Harris_2012]. In contrast to recent years, between 1995 and 2006, an average of `r mean(time_data[as.character(1995:2006), 'b_n50'])` studies were responsible for submitting more than half of the sequences each year. Although these deep surveys represent significant contributions to our knowledge of bacterial biogeography, their small number and lack of environmental diversity is indicative of the broader problems in advancing the bacterial census.

***The dynamics of the archaeal census.***
The depth of sequencing being done to advance the archaeal census is `r floor(sum(time_data$b_nseqs)/sum(time_data$a_nseqs))`-times less than that of the bacterial census. In contrast to the plateau that was observed in bacterial sequence submissions, prior to 2012 the submission of archaeal sequences generally increased each year; however between 2012 and 2014 the number of archaeal sequences submitted decreased from `r time_data['2012', 'a_nseqs']` to `r time_data['2014', 'a_nseqs']` sequences per year. The number of new OTUs contributed each year has largely paralleled that of the bacterial census with a plateau starting in 2005 at `r mean(time_data[as.character(2005:2014),'a_sobs'])` new OTUs per year. Although the number of submissions responsible for more than 50% of the archaeal sequences submitted per year has never been more than `r max(time_data[,'a_n50'])`, with the exception of 2012 and 2014 there has been an average of `r mean(time_data[as.character(c(1996:2011,2013)),'a_n50'])` submissions responsible for more than 50% of the sequences each year since 1996. The clear bias towards sequencing bacterial 16S rRNA genes has limited the ability to more fully characterize the biodiversity of the archaea.

* What can we do to fix this?
  - Characterize coverage by environments
  - Cultured vs. uncultured
  - Single cell genomics

@@

```{r figure4a, eval=FALSE}
data <- read.table(file="data/mothur/all_bacteria.env_category.groups.rarefaction", header=T)
category <- "\\.plant$" #change the SO to whatever category you want (leave the \\. and $)
plant_columns <- grep(category, colnames(data))
category <- "\\.aerosol$" #change the SO to whatever category you want (leave the \\. and $)
aerosol_columns <- grep(category, colnames(data))
category <- "\\.aquatic$" #change the SO to whatever category you want (leave the \\. and $)
aquatic_columns <- grep(category, colnames(data))
category <- "\\.built$" #change the SO to whatever category you want (leave the \\. and $)
built_columns <- grep(category, colnames(data))
category <- "\\.other$" #change the SO to whatever category you want (leave the \\. and $)
other_columns <- grep(category, colnames(data))
category <- "\\.soil$" #change the SO to whatever category you want (leave the \\. and $)
soil_columns <- grep(category, colnames(data))
category <- "\\.zoonotic$" #change the SO to whatever category you want (leave the \\. and $)
zoonotic_columns <- grep(category, colnames(data))
plant_aerosol_aquatic_built_other_soil_zoonotic_data <- data[,c(1, plant_columns, aerosol_columns, aquatic_columns, built_columns, other_columns, soil_columns, zoonotic_columns)]  #the 1 gets the first column which is numsampled
nrows_to_keep <- which.max(plant_aerosol_aquatic_built_other_soil_zoonotic_data$numsampled[!is.na(plant_aerosol_aquatic_built_other_soil_zoonotic_data[,2])])
plant_aerosol_aquatic_built_other_soil_zoonotic_data <- plant_aerosol_aquatic_built_other_soil_zoonotic_data[1:nrows_to_keep,]
pdf(file="/Users/renegirard7gfy/Desktop/Bac_Uniq_3per.pdf")

@@

```{r figure4b, eval=FALSE}
